Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
working on #33

# Abstract (150 words) {-}
Increased frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
Tree species adaptations to hot and dry conditions -- e.g., dry-season deciduousness in tropical seasonal forests -- can confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
While individuals with more exposed crowns tended to grow less in all droughts, tree size and topographic wetness had variable effects.
<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, <!--tree size--> and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location,
iii) the influences of topographic location, crown exposure, and other size effects on individual drought responses are variable across droughts.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.

@@

<!---CONCLUSIONS---->
We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), severe drought in the wet-season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`). 
<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
Divergent effects of species and individuals to droughts have  may contribute to the maintenance of species diversity and ecosystem stability.

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potnetially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

>>>>

# Abstract (150 words) {-}
Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
Tree species adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
While individuals with more exposed crowns tended to grow less in all droughts, stem diameter and topographic wetness had variable effects.
<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if these factors act independently, tree drought responses to may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought in one particular environment may prove vulnerable to a drought with different characteristics or in a different environment. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) on the species level, deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location;
iii) on the individual level, the influences of topographic location, crown exposure, and other size effects on drought responses are variable across droughts.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.

@@

<!---CONCLUSIONS---->
We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), severe drought in the wet-season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`). 
<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
Divergent effects of species and individuals to droughts have  may contribute to the maintenance of species diversity and ecosystem stability.

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->
